Combinatorics of the geometry of Wilson loop diagrams II: Grassmann necklaces, dimensions, and denominators

### 2019-08-15 GWLDCII.tex
some edits and comments up to p23

Let $[n]$ denote the set of integers $\{1,\dots,n\}$ with the obvious cyclic and linear orders. Our primary focus is Wilson loop diagrams with $V = [n]$, so in order to simplify the exposition we will often use $i+1$ as shorthand for ``the successor of vertex $i$''.

@@

In particular, we will see in later sections that all four vertices supporting a propagator are important in understanding the contribution of that propagator to the \st{overall diagram}{geomtric object represented by the diagram} \sanote{Is this what we mean?}. This motivates the following definition of a subdiagram, any subset $Q$ of the propagators and any subset $U$ of the vertices of $W$ {\em containing the support of $Q$}.

@@

These diagrams correspond to (tree level) amplitudes of SYM $N=4$ theory in twistor space. As such, it associates a real value to $n$ particles, one for each vertex. Each particle is represented as a section of a $|\cP|$-vector bundle over twistor space, projected onto a real subspace in a process called bosonization \cite{Arkani-Hamed:2013jha}. In other words, each Wilson loop diagram defines a functional on external particles. This is represented as an integral $I(W)$, which is the equivalent of a Feynman integral in this setting. \sanote{This feels like too much detail. Thoughts?} This integral is comprised of two main components, and interaction matrix, $C(W)$ and a volume form, $\Omega(W)$. These are described below.

@@

The matrix $C(W)$ can be interpreted as a \hlfix{parametrization of the positroid cell associated to $W$}{I know we've been saying this, but we have $4k$ parameters for a $3k$ dimensional space. Is this okay?} in the positive Grassmannian $\Gr(n, |\cP|)$, which we denote by $\Sigma(W)$ \cite{wilsonloop}. Most of the work in this paper, and in \cite{generalcombinatoricsI} focuses on characterizing the positroid cell associated to each Wilson loop diagram $W$.

@@

\begin{dfn}\label{def R(W)}
Let $W = (\cP,V)$ be an admissible Wilson loop diagram, and fix an edge $e$. Let $\{p_1 \ldots p_r\}$ be the propagators with one end lying on edge $e$ under a counterclockwise orderin. That is, $p_1$ is closest to the vertex $e$, $p_i$ is positioned counterclockwise to $p_{i+1}$ and $p_r$ is the clockwise most propagator on this edge (closest to $e+1$). Define the polynomial
\[ R_e =  c_{p_1,e+1} \prod_{j= 1}^{r-1} \left((c_{p_j,e} c_{p_{j+1},e+1} - c_{p_{j+1},e} c_{p_{j},e+1} ) \right) c_{p_r,e}\;\]
to be the component of $R(W)$ associated to edge $e$. Note that if $r = 1$ (i.e. there is exactly one propagator lying on edge $e$) then this expression simplifies to $R_e = c_{p,e} c_{p,e+1}$. If $r=0$, set $R_e = 1$. The denominator $R(W)$ is defined to be the product of all of the $R_e$, i.e. 
\[R(W) = \prod_{e \in V} R_e.\]
\end{dfn}

@@

For our purposes, a {\bf positroid} is a matroid $M$ (with ground set $[n]$ and bases $\cB \subseteq \binom{[n]}{k}$) which can be represented by an element of the positive Grassmannian $\Gr(k,n)$. In other words, there exists a full-rank $k\times n$ real matrix whose $k\times k$ minors are all nonnegative and such that the minor $[[k]|J]$ is positive if and only if $J \in \cB$.

@@

For each $j \in [n]$, we can define a total order $<_j$ on the interval $[n]$ by
\[ j <_j j+1 <_j \dots <_j n <_j 1 \dots <_j j-1\;.\]
This in turn induces a total order on $\binom{[n]}{k}$, namely the lexicographic order with respect to $<_j$.  It also induces a separate partial order $\gale{j}$ on $\binom{[n]}{k}$ (the {\bf Gale order} \cite{Gale}), which is defined as follows: for 
\[A = \{a_1 <_j a_2 <_j \dots <_j a_k\} \text{ and } B = \{b_1 <_j b_2 <_j \dots <_j b_k\} \in \binom{[n]}{k},\] we define
\[A \gale{j} B \text{ if and only if } a_r \leq_j b_r \text{ for all }1 \leq r \leq k.\]
For example, in $\binom{[6]}{3}$ we have $\{2,5,6\}\gale{2} \{2,6,1\}$ but $\{2,5,6\}\not\gale{2}\{3,4,6\}$.

@@

In \cite{wilsonloop}, Agarwala and Marin-Amat show that $M(W)$ is also characterized by the Wilson loop diagram itself:

@@

\begin{dfn}\label{props inside p}
Let $W = (\cP, [n])$ be a weakly admissible Wilson loop diagram, with $p \in \cP$ supported on edges $i$ and $j$. Write $(p, <_i)$ to represent the same propagator directed from $i$ to $j$. Then the sets of propagators {\bf inside} \sanote{just checking: $p \in \cP_{in}(p)$?} and {\bf outside} of $(p,<_i)$ are defined to be
\begin{gather*}\cP_{in}(p,<_i) = \{ (k,l) \in \cP \ |\ i \leq_i k <_i l \leq_i j \}, \\
\cP_{out}(p,<_i) = \cP\setminus \cP_{in}(p,<_i),
\end{gather*}
respectively.
\end{dfn}

@@

\begin{lem}\label{lem sian}
  Let $W$ be a weakly admissible WLD with at least 5 vertices and in which each vertex supports at least one propagator.  Then at least one of the following two things occurs.
  \begin{enumerate}
    \item $W$ has a propagator of length $\leq 6$ with a propagator of length 2 on one side it and nothing else on that side.\label{item big and 2}
    \item There exists a pair of propagators of length $2$ with the property that the first propagator is $(i, i+2)$, the second is $(j, j+2)$, no other propagator ends between vertices $i+2$ and $j+1$, and $j\in\{i+2, i+3, i+4\}$.\label{item pair of 2s}
  \end{enumerate}
\end{lem}

@@

Let \hlfix{$W_r := (\cP_{in}(q_r),[n])$}{technically has non supporting vertices. Want $(\cP_{in}(q_r),[i_r, k_r+1])$. However this is clear later. Do we care?} be the subdiagram of $W$ consisting only of those propagators inside $q_r$ (including $q_r$ itself); in particular $W_r$ contains none of the previous $p_i$ or $q_i$. By the original hypotheses on $W$, every vertex in $[i_r,k_{r+1}]$ (of which there are at least 4, since $\ell(q_r) \geq 4$) must support at least one propagator. Therefore $W_r$ admits at least one propagator of length 2 (by Remark~\ref{rem:props of length 2 and 3}) and no propagators of length $3$ (by our assumption on $W$).

@@

Note that we could replace $q_r$ by another propagator $q'_{r}$ in \st{$W_r$} {\color{violet}$W_r = (\cP_{in}(q_r), [n])$}\sanote{just want to emphasize that $q_r'$ is inside $q_r$}, as long as $p_{r+1}$ and $q_r$ are on opposite sides of $q_r'$: such a new $q'_{r}$ would still satisfy all of the induction hypotheses. (See Figure~\ref{fig smallest q in lem sian}.) Therefore without loss of generality we may assume that $\cP_{in}(q_r)$ does not contain any propagators (other than $q_r$ itself) which satisfy the induction hypotheses, \st{i.e. we assume that there are no other propagators with $p_{r+1}$ on one side and $q_r$ on the other}.

@@

@@

The second case, \hlfix{there are at least $3$ vertices between one endpoint of $p_{r+1}$ and the nearest endpoint of $q_r$.}{Does this just convolute matters?} In other words,  (at least) one of the following four things happen:
\begin{itemize} 
\item $i_r+3 \leq j_{r+1}\leq k_r-3$ (one endpoint of $p_{r+1}$ has at least 3 vertices between it and the nearest enpoint of $q_r$), or
\item $i_r+3 \leq j_{r+1}+2\leq k_r-3$ (one endpoint of $p_{r+1}$ has at least 3 vertices between it and the nearest enpoint of $q_r$), or
\item $i_r = j_{r+1}$ ($p_{r+1}$ and $q_r$ share and end edge), or 
\item $k_r = j_{r+1} +2$  ($p_{r+1}$ and $q_r$ share and end edge)
\end{itemize}
In other words, at least one end of $p_{r+1}$ is supported entirely by vertices in the interval $[i_r+3,k_r+2]$ (the first two bullet points) or both ends lie in $[i_r, i_r+3]$ (third point) or both ends lie in $[k_r-2, k_r+1]$ (fourth point). These situations all behave similarly, and by symmetry it suffices to only consider the second and third possibilities.

@@

The overall result then follows by induction.
\end{proof}

@@

Until now, the positroid associated to a Wilson loop diagram $W$ could only be obtained by computing the matrix $C(W)$ associated to $W$, listing all bases of the induced matroid $M(W)$, and constructing the Le diagram or Grassmann necklace of the positroid ``by eye'' from this list.

@@

\begin{algorithm}\label{alg:put GN on WLD}
Let $W = (\cP, [n])$ be an admissible Wilson loop diagram. In order to construct the set set $I_i^W$ (or just $I_i$ when $W$ is clear from context) for $i \in [n]$, perform the following steps:

@@

To construct $I_1$ we start at vertex 1, so we set $i=1$, $j = 1$, and $I_1 = \emptyset$. 
\begin{itemize}
\item $j = 1$: Since $\Prop(1) = \{q,r\}$ and $r$ is the clockwise-most of these propagators, we let $I_1 = \{1\}$ and remove propagator $r$ from the diagram. 
\item $j = 2$: $\Prop(2) = \{p,q\}$, with $q$ being the clockwise-most of these propagators. So $I_1 = \{1,2\}$ and propagator $q$ is removed.
\item $j = 3$: $\Prop(3) = \{p\}$, so now $I_1 = \{1,2,3\}$ and propagator $p$ is removed. The only remaining propagator in the diagram is $s$.
\item $j = 4$: Since $p$ and $q$ were removed in earlier steps, we now have $\Prop(4) = \emptyset$.
\item $j = 5$: $\Prop(5) = \{s\}$, and we have $I_1 = \{1,2,3,5\}$.
\end{itemize}
There are no propagators left in the diagram, so the algorithm terminates and we have ${I_1 = \{1,2,3,5\}}$, or $1235$ for short. 
%So the propagators are ordered $r,q,p,s$ with respect to $I_1$, and 
$I_1$ viewed as a function is given by
\[I_1: \cP \longrightarrow [n] : \quad I_1(r) = 1,\quad I_1(q) = 2, \quad I_1(p) = 3, \quad I_1(s) = 5.\]
Applying the algorithm for all 8 starting vertices, we obtain the sets
\[1235, 2356, 3456, 4567, 5671, 6712, 7812, 8123.\]
The reader can easily verify that this sequence of $k$-sets satisfies the definition of a Grassmann necklace, and can (with significantly more work) also verify that this Grassmann necklace defines the positroid $M(W)$ associated to the Wilson loop diagram $W$ above.
\end{eg}

@@

We now consider what happens as we move from $V$ to each of the three remaining cases for $W$. We need to consider both $J^{W}_p$ and $J^{W}_r$ for $r \neq p$, since the addition of $p$ can have a knock-on effect on later steps in the algorithm.

>>>>

Let $[n]$ denote the set of integers $\{1,\dots,n\}$ with the obvious cyclic and linear orders. Our primary focus is Wilson loop diagrams with $V = [n]$, so in order to simplify the exposition we will often use $i+1$ as shorthand for ``the successor of vertex $i$''. \note{We say this here and then again in the paragraph after length.  Can we discuss how to say it once in a clear and concise way.}

@@

In particular, we will see in later sections that all four vertices supporting a propagator are important in understanding the contribution of that propagator to the \st{overall diagram}{geometric object represented by the diagram} \sanote{Is this what we mean?}\note{Is this chatty bit even helpful at all: this seems like the most natural notion of subobject anyway.}. This motivates the following definition of a subdiagram, any subset $Q$ of the propagators and any subset $U$ of the vertices of $W$ {\em containing the support of $Q$}.

@@

These diagrams correspond to (tree level) amplitudes of SYM $N=4$ theory in twistor space. As such, it\note{What is this ``it'', is it ``these diagrams'' if so there's a singular/plural agreement problem, if not then it isn't clearly enough written.} associates a real value to $n$ particles, one for each vertex. Each particle is represented as a section of a $|\cP|$-vector bundle over twistor space, projected onto a real subspace in a process called bosonization \cite{Arkani-Hamed:2013jha}. In other words, each Wilson loop diagram defines a functional on external particles. This is represented as an integral $I(W)$, which is the equivalent of a Feynman integral in this setting. \sanote{This feels like too much detail. Thoughts?}\note{Seems ok to me, at least with the grammar fixed as noted above} This integral is comprised of two main components, and interaction matrix, $C(W)$ and a volume form, $\Omega(W)$. These are described below.

@@

The matrix $C(W)$ can be interpreted as a \hlfix{parametrization of the positroid cell associated to $W$}{I know we've been saying this, but we have $4k$ parameters for a $3k$ dimensional space. Is this okay?}\note{Maybe there's a better word than ``parametrization''?  It does define the positroid cell by giving its points, we wouldn't need to say parametrizaion to say that.} in the positive Grassmannian $\Gr(n, |\cP|)$, which we denote by $\Sigma(W)$ \cite{wilsonloop}. Most of the work in this paper, and in \cite{generalcombinatoricsI} focuses on characterizing the positroid cell associated to each Wilson loop diagram $W$.

@@

\begin{dfn}\label{def R(W)}
Let $W = (\cP,V)$ be an admissible Wilson loop diagram, and fix an edge $e$. Let $\{p_1 \ldots p_r\}$ be the propagators with one end lying on edge $e$ under a counterclockwise orderin. That is, $p_1$ is closest to the vertex $e$, $p_i$ is positioned counterclockwise to $p_{i+1}$ and $p_r$ is the clockwise most propagator on this edge (closest to vertex $e+1$). Define the polynomial
\[ R_e =  c_{p_1,e+1} \prod_{j= 1}^{r-1} \left((c_{p_j,e} c_{p_{j+1},e+1} - c_{p_{j+1},e} c_{p_{j},e+1} ) \right) c_{p_r,e}\;\]
to be the component of $R(W)$ associated to edge $e$. Note that if $r = 1$ (i.e. there is exactly one propagator lying on edge $e$) then this expression simplifies to $R_e = c_{p,e} c_{p,e+1}$. If $r=0$, set $R_e = 1$. The denominator $R(W)$ is defined to be the product of all of the $R_e$, i.e. 
\[R(W) = \prod_{e \in V} R_e.\]
\end{dfn}

@@

For our purposes, a {\bf positroid} is\footnote{Note that matroid isomorphism allows arbitrary permutations of the ground set but positroid isomorphism allows only cyclic permutations, so the property of being a positroid is not in general preserved by matroid isomorphism.} a matroid $M$ (with ground set $[n]$ and bases $\cB \subseteq \binom{[n]}{k}$) which can be represented by an element of the positive Grassmannian $\Gr(k,n)$. In other words, there exists a full-rank $k\times n$ real matrix whose $k\times k$ minors are all nonnegative and such that the minor $[[k]|J]$\note{I don't think we have this $[[k]|J]$ notation defined?} is positive if and only if $J \in \cB$.

@@

For each $j \in [n]$, we can define a total order $<_j$ on the interval $[n]$ by
\[ j <_j j+1 <_j \dots <_j n <_j 1 \dots <_j j-1\;.\]
This in turn induces a total order on $\binom{[n]}{k}$, namely the lexicographic order with respect to $<_j$.  It also induces a separate partial order $\gale{j}$ on $\binom{[n]}{k}$ (the {\bf Gale order} \cite{Gale}\note{We only use this once and in this particular section, so maybe we don't need to define and name it but can just put the def in-line where we need it}), which is defined as follows: for 
\[A = \{a_1 <_j a_2 <_j \dots <_j a_k\} \text{ and } B = \{b_1 <_j b_2 <_j \dots <_j b_k\} \in \binom{[n]}{k},\] we define
\[A \gale{j} B \text{ if and only if } a_r \leq_j b_r \text{ for all }1 \leq r \leq k.\]
For example, in $\binom{[6]}{3}$ we have $\{2,5,6\}\gale{2} \{2,6,1\}$ but $\{2,5,6\}\not\gale{2}\{3,4,6\}$.

@@

In \cite{wilsonloop}, Agarwala and Marin-Amat show that $M(W)$ is also characterized by the Wilson loop diagram itself\note{I don't like this phrasing; can we discuss}:

@@

\begin{dfn}\label{props inside p}
Let $W = (\cP, [n])$ be a weakly admissible Wilson loop diagram, with $p \in \cP$ supported on edges $i$ and $j$. Write $(p, <_i)$ to represent the same propagator directed from $i$ to $j$. Then the sets of propagators {\bf inside} \sanote{just checking: $p \in \cP_{in}(p)$?} and {\bf outside} of $(p,<_i)$ are defined to be
\begin{gather*}\cP_{in}(p,<_i) = \{ (k,l) \in \cP \ |\ i \leq_i k <_i l \leq_i j \}, \\
\cP_{out}(p,<_i) = \cP\setminus \cP_{in}(p,<_i),
\end{gather*}
respectively.
\end{dfn}
Note that we have $p\in \cP_{in}(p, <_i)$ and so $\cP_{in}(p, <_i) = \cP_{out}(p, <j)\cup\{p\}$.

@@

\begin{lem}\label{lem sian}
  Let $W$ be a weakly admissible WLD with at least 5 vertices and in which each vertex supports at least one propagator.  Then at least one of the following two things occurs.
  \begin{enumerate}
    \item $W$ has a propagator of length $\leq 6$ with a propagator of length 2 on one side of it and nothing else on that side.\label{item big and 2}
    \item There exists a pair of propagators of length $2$ with the property that the first propagator is $(i, i+2)$, the second is $(j, j+2)$, no other propagator ends between vertices $i+2$ and $j+1$, and $j\in\{i+2, i+3, i+4\}$.\label{item pair of 2s}
  \end{enumerate}
\end{lem}
\note{Should this lemma have pictures}

@@

Let \hlfix{$W_r := (\cP_{in}(q_r),[n])$}{technically has non supporting vertices. Want $(\cP_{in}(q_r),[i_r, k_r+1])$. However this is clear later. Do we care?}\note{No, we don't quite want that either because we don't want the newly shortened $q_r$ to be our propagator of length 2 -- maybe a slight change to the remark} be the subdiagram of $W$ consisting only of those propagators inside $q_r$ (including $q_r$ itself); in particular $W_r$ contains none of the previous $p_i$ or $q_i$. By the original hypotheses on $W$, every vertex in $[i_r,k_{r+1}]$ (of which there are at least 4, since $\ell(q_r) \geq 4$) must support at least one propagator. Therefore $W_r$ admits at least one propagator of length 2 (by Remark~\ref{rem:props of length 2 and 3}) and no propagators of length $3$ (by our assumption on $W$).

@@

Note that we could replace $q_r$ by another propagator $q'_{r}$ in \st{$W_r$} {\color{violet}$W_r = (\cP_{in}(q_r), [n])$}\sanote{just want to emphasize that $q_r'$ is inside $q_r$}, as long as $p_{r+1}$ and $q_r$ are on opposite sides of $q_r'$: such a new $q'_{r}$ would still satisfy all of the induction hypotheses. (See Figure~\ref{fig smallest q in lem sian}.) Therefore without loss of generality we may assume that $\cP_{in}(q_r)$ does not contain any propagators (other than $q_r$ itself) which satisfy the induction hypotheses, \st{i.e. we assume that there are no other propagators with $p_{r+1}$ on one side and $q_r$ on the other}.\note{I kind of like the struck part, can we make your other changes and also keep the struck part.}

@@

\note{Would it be worth having vertex $j_{r+1}$ labelled in this fig smallest q in lem sian}

@@

The second case, \hlfix{there are at least $3$ vertices between one endpoint of $p_{r+1}$ and the nearest endpoint of $q_r$.}{Does this just convolute matters?}\note{Well at some stage it got edited so that it isn't a sentence any longer, but I think something like this should be here because it defines what the second case is.} In other words,  (at least) one of the following four things happen:
\begin{itemize} 
\item $i_r+3 \leq j_{r+1}\leq k_r-3$ (one endpoint of $p_{r+1}$ has at least 3 vertices between it and the nearest enpoint of $q_r$), or
\item $i_r+3 \leq j_{r+1}+2\leq k_r-3$ (one endpoint of $p_{r+1}$ has at least 3 vertices between it and the nearest enpoint of $q_r$), or
\item $i_r = j_{r+1}$ ($p_{r+1}$ and $q_r$ share and end edge), or 
\item $k_r = j_{r+1} +2$  ($p_{r+1}$ and $q_r$ share and end edge)
\end{itemize}
In other words, at least one end of $p_{r+1}$ is supported entirely by vertices in the interval $[i_r+3,k_r+2]$ (the first two bullet points) or both ends lie in $[i_r, i_r+3]$ (third point) or both ends lie in $[k_r-2, k_r+1]$ (fourth point). These situations all behave similarly, and by symmetry it suffices to only consider the second and third possibilities.

@@

The overall result then follows by induction.

\end{proof}

@@

Until now, the positroid associated to a Wilson loop diagram $W$ could only be obtained by computing the matrix $C(W)$ associated to $W$, listing all bases of the induced matroid $M(W)$, and constructing the Le diagram or Grassmann necklace of the positroid ``by eye'' from this list. \note{Can we discuss how to say this in a way that doesn't appear weak.}

@@

\begin{algorithm}\label{alg:put GN on WLD}
Let $W = (\cP, [n])$ be an admissible Wilson loop diagram. In order to construct the set set $I_i^W$ (or just $I_i$ when $W$ is clear from context\note{We have gone from not pointing out this convention to pointing it out 3 times. Can we find a happy medium?}) for $i \in [n]$, perform the following steps:

@@

To construct $I_1$ we start at vertex 1, so we set $i=1$, $j = 1$, and $I_1 = \emptyset$. 
\begin{itemize}
\item $j = 1$: Since $\Prop(1) = \{q,r\}$ and $r$ is the clockwise-most of these propagators, we let $I_1 = \{1\}$ and remove propagator $r$ from the diagram. 
\item $j = 2$: $\Prop(2) = \{p,q\}$, with $q$ being the clockwise-most of these propagators. So $I_1 = \{1,2\}$ and propagator $q$ is removed.
\item $j = 3$: $\Prop(3) = \{p\}$, so now $I_1 = \{1,2,3\}$ and propagator $p$ is removed. The only remaining propagator in the diagram is $s$.
\item $j = 4$: Since $p$ and $q$ were removed in earlier steps, we now have $\Prop(4) = \emptyset$.
\item $j = 5$: $\Prop(5) = \{s\}$, and we have $I_1 = \{1,2,3,5\}$.
\end{itemize}
There are no propagators left in the diagram, so the algorithm terminates and we have ${I_1 = \{1,2,3,5\}}$, or $1235$ for short. 
%So the propagators are ordered $r,q,p,s$ with respect to $I_1$, and 
$I_1$ viewed as a function is given by
\[I_1: \cP \longrightarrow [n] : \quad I_1(r) = 1,\quad I_1(q) = 2, \quad I_1(p) = 3, \quad I_1(s) = 5.\]
Applying the algorithm for all 8 starting vertices, we obtain the sets
\[1235, 2356, 3456, 4567, 5671, 6712, 7812, 8123.\]
The reader can easily verify that this sequence of $k$-sets satisfies the definition of a Grassmann necklace, and can (with significantly more work) also verify that this Grassmann necklace defines the positroid $M(W)$ associated to the Wilson loop diagram $W$ above, as we will prove in general shortly.
\end{eg}

@@

We now consider what happens as we move from $V$ to each of the three remaining cases for $W$. We need to consider both $J^{W}_p$ and $J^{W}_r$ for $r \neq p$, since the addition of $p$ can have a knock-on\note{Wouldn't the expression be knock off?} effect on later steps in the algorithm.
